---
title: "final"
format: html
---

https://github.com/misssammwoods04/ENVS-193DS_spring-2025_final

```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
library(here)
library(DHARMa)
library(ggeffects)
library(zoo)
library(MuMIn)
library(scales)
library(readxl)
```


# Problem 1

## a.

In part one, one co-worker used a linear regression test. In part two, they used and analysis of variance test (ANOVA). 

## b. 

The F distribution should be included to provide more context because it would help us determine significance, which helps us assess whether the observed differences are likely due to an actual relationship or random variation. since ANOVA only tells us that at least one group differs from the others, it would be important to conduct a post-hoc test (such as Tukey’s HSD) to identify which specific groups differ significantly from each other.

## c.

We found a strong correlation between annual total nitrogen laod nitrogen load (kg/year) and distance from the headwater (km) (Linear regression, r value, F distribution, p = 0.03, $\alpha$ = significance level)

Average nitrogen load (kg year⁻¹) differed significantly among nitrogen sources. Post-hoc testing confirmed significant differences between certain categories (One-way ANOVA, F distribution, Tukey HSD post-hoc test, p = 0.02, $\alpha$ = significance level)


# Problem 2

```{r}
sst <- read.csv("data/SBH_temperature_20241231.csv")
```


## a.

```{r}

sst_clean <- sst |> # data frame
  clean_names() |> # clean names
   mutate(month = factor(month.name[month], levels = month.name)) |>
  select(year, month, surf_temp_c) |> # only show these three columns
  filter (year %in% c(2018:2023)) |> # only show these years
  group_by(year, month) |> # groups data by year and month
  summarise(mean_monthly_sst = mean(surf_temp_c, na.rm = TRUE),  .groups = "drop") |> # find mean sst of each month based on year
  mutate(mean_monthly_sst = zoo::na.approx(mean_monthly_sst, na.rm = FALSE)) # change month number to name (abbreviate)

```


```{r}
sst_clean |> 
  slice_sample(n = 5) # Display 5 random rows from sst_clean

str(sst_clean)# Show the structure of sst_clean
```

```{r}
ggplot(sst_clean, 
       aes(x = month, # x axis
           y = mean_monthly_sst, # y-axis
           color = factor(year), # color based on year
           group = year)) + # group based on year
  geom_point() + # include points for each observation
  geom_line() + # connect points
  coord_cartesian(ylim = c(12, 20)) +   # Zoom y-axis to 14-20, keep all data
  scale_y_continuous(
    breaks = seq(14, 20, by = 2), # determines range of y axis 
    labels = c (14, 16, 18, 20) # determines what shows up on y axis
  ) +
  labs(x = "Month", # x axis name
       y = "Mean monthly sea surface temperature (\u00B0C)", # y axis name
       color = "Year") + # color based on year
  theme_minimal() + # theme
   theme(
    panel.grid.major = element_blank(),  # remove major grid lines
    panel.grid.minor = element_blank()   # remove minor grid lines
  ) +
  theme(panel.border = # outline grid
          element_rect(color = "black", # color outline in black
                       fill = NA, size = 0.5), # thickness of outline
        axis.ticks.x = element_line(), # add x axis ticks
        legend.position = c(0.1, 0.8)) # change legend position
       
```


# Problem 3
 

```{r}
nest_boxes <- read_csv("data/occdist.csv")

clean_nest_boxes <- nest_boxes |> 
  clean_names() |> 
  select(sp, edge_distance, season, box_occupant) |> 
mutate(box_occupant = as_factor(box_occupant))
```

## a. 

The 1s indicate the presence of something (yes) and the 0s indicate the absence of something (no). 

## b.

The main difference between Swift Parrots and the other two species is that Swift Parrots are the target species of the nest boxes while the other species are not.

## c. 

The two years are 2016 and 2019, and they differ based on the target and non-target species occupancy. 

## d.

| Model number | E_D | Season | Predictor list             |
|:------------:|:---:|:------:|----------------------------|
| 0            |     |        | no predictors (null model) |
| 1            |  X  |   X    | all predictors (full model)| 
| 2            |  X  |        | ED                         |   
| 3            |     |   X    | season                     |

## e.

```{r}
model0 <- lm(
  sp ~ 1,
  data = clean_nest_boxes
)

model1 <- lm(
  sp ~ edge_distance + season, 
  data = clean_nest_boxes
)

model2 <- lm(
  sp ~ edge_distance ,
  data = clean_nest_boxes
)

model3 <- lm(
  sp ~ season ,
  data = clean_nest_boxes
)
```

## f. 

```{r}

par(mfrow = c(2,2))
plot(model0)
plot(model1)
plot(model2)
plot(model3)

```


## g. 

```{r}

AICc(
  model0,
  model1,
  model2,
  model3
) |> 
  arrange(AICc)

```

```{r}
summary(model1)
```

The best model that explains total plant math includes edge distance and season (F(2,224 = 8.5, p < 0.001, $\aplha$ = 0.05).

## h. 

```{r}
model1_predictions <- ggpredict(
  model1, # model object
  terms = c("edge_distance", "season") # predictors
) |> 
  # treating this like a regular data frame
  # renaming the columns
  rename(edge_distance = x,
         season = group)
```


```{r}

clean_nest_boxes$season <- factor(clean_nest_boxes$season, levels = c("2016", "2019")) # show seasons as separate data sets 

model1_predictions$season <- factor(model1_predictions$season, levels = c("2016", "2019")) # show seasons as separate data sets 
 
ggplot(data = clean_nest_boxes, # data set
       aes(x = season, # x-axis = season
           y = edge_distance, # y-axis = edge distance
           color = season )) + # color based on season
  
  #control jitter
  geom_jitter(width = 0.1,
              height = 0,
              shape = 21,
              alpha = 0.6) +
  
  facet_wrap(~box_occupant, nrow = 1) + # makes each panel represent data fro the different species
  
  labs(x = "Season", # change x-axis name
       y = "Edge Distance") + # change y-axis name
  
  scale_color_manual(values = c("2016" = "#1b9e07", "2019" = "#d95f92")) + # custom colors
  
  stat_summary(geom = "point", # code a circle that represents the mean
               fun = mean, # circle represents mean
               size = 4) + # make circle larger than observations points
  
  theme(legend.position = "none", # remove legend
         panel.grid.major = element_blank(),  # remove major grid lines
    panel.grid.minor = element_blank()   # remove minor grid lines
        )
  
```

## i.

Figure 1. Non target species (common starling, tree martin) tend to occupy nest boxes. 
Data from "occdist.csv" file (Stojanovic D, Owens G, Young C, Alves F, 2021).
Points represent observations of nest box occupation by target and non target species (n=227). Colors represent nest box occupation based on season (2016 or 2019). Large, filled in circles reprsent the mean of each season of observations. 

j. 

```{r}

n_mod <- glm(sp ~ edge_distance, # creates model for linear regression
             data = clean_nest_boxes, # data frame
             family = "binomial") # identifies model a binomial

# Simulate and plot residuals
plot(
  simulateResiduals(n_mod) 
)

summary(n_mod) # gives summary of logistic regression model

gtsummary::tbl_regression(n_mod,
                          exponentiate = TRUE) # exponentiates logistic regression model 

exp(coef(n_mod))  # exact odds ratio for edge_distance

ggpredict (n_mod,
           terms = "edge_distance [0]") # gives predicted probability for edge distance = 0 

ggpredict(n_mod,
          terms = "edge_distance [300]") # gives predicted probability for edge distance = 300

ggpredict(n_mod,
          terms = "edge_distance [600]") # gives predicted probability for edge distance = 600

ggpredict(n_mod,
          terms = "edge_distance [900]") # gives predicted probability for edge distance = 900

```


## k.

With every 1 m increase in distance from the forest edge, the odds of a Swift Parrot occupying a nest box slightly decrease, although the effect is very small (odds ratio = 0.998, 95% CI [1.00, 1.00], p = 0.001, α = 0.05). While this odds ratio is very close to 1, the predicted probability of occupancy still decreases as distance increases.

This trend suggests that Swift Parrots are more likely to occupy nest boxes located closer to the forest edge. One possible biological explanation is that non-target species may dominate boxes farther from the edge, limiting Swift Parrot occupancy in those areas.

# Problem 4


My homework 3 visualization is a lot more detailed and creative and includes all my predictor variables while my initial visualization shows only one predictor variable. Also, my initial visualization is more straightforward because it is a simple graph while my final visualization is more complex. 

Between all visualizations I see a similarity in predictor variables as I kept the same predictors.

I don't see many patterns throughout my visualizations, especially because my initial idea included on a graph and my final visualization was a lot more creative than I initially intended. At first, I was focused on making a straight forward, logical visualization, but as I realized that I could  show my data however I wanted, I leaned towards a more creative vision.

The main feedback I received was that there was not much indication of a correlation between my predictor and response variables. I knew that there was not a correlation between my variables, but I decided to show this by running a logisitic regression on my website. This test confirmed my assumption that there was no correlation between my predictors and response variables. 
